### 2013-04-08 inst/doc/beyond-mechanistic-models.Rmd
writing writing

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). "}
```
```{r policies_plot, dependson=c("opt", "plotting-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979]."
}
```
```{r sim_plot, dependson=c("sim", "plotting-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."
}
```

@@

The dynamics occur over a range of state-space in which the biologically
implausible model performs as well or better than the more plausible
model.  This is the most obvious and immediate reason why the shortcomings
that appear to make the model implausible do not make it useless.
This effect is enhanced in our example because the management actions help
drive the system towards rather than away from this region of state-space.

>>>>

```{r gp_plot, dependson=c("gp", "mle", "plotting-options"), fig.cap="Graph of the inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  Graph shows the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data is also shown as black points.  (The GP is conditioned on 0,0, shown as a pseudo-data point). "}
```

```{r policies_plot, dependson=c("opt", "plotting-options"), fig.cap="The steady-state optimal policy (infinite boundary) calculated under each model.  Policies are shown in terms of target escapement, $S_t$, as under models such as this a constant escapement policy is expected to be optimal [@Reed1979]."}
```

```{r sim_plot, dependson=c("sim", "plotting-options"), fig.cap="Gaussian process inference outperforms parametric estimates. Shown are 100 replicate simulations of the stock dynamics (eq 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model."}
```

@@

The dynamics occur over a range of state-space in which the biologically
implausible model performs as well or better than the more plausible
model.

This is the most obvious and immediate reason why the shortcomings
that appear to make the model implausible do not make it useless.
This effect is enhanced in our example because the management actions help
drive the system towards rather than away from this region of state-space.

### 2013-04-08 inst/doc/nonparametric-bayes.tex

 Okay, perhaps that's novel, but it's pretty trivial. Isn't it obvious
to everyon that it's trivial?

 paraphrase advantages of machine learning

@@

The essense of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.\\
Figure 1 include curves drawn from the posterior density?

 Do we need more specifics on Gaussian process as an approximation to
parametric models? Discussion of Gaussian process vs other machine
learning / forecasting approaches that have less clear statistical
foundations? If so, does this all belong in the discussion?

@@

@@

\subsubsection{The optimal control problem}

@@

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). To compare
the performance of nonparametric and parametric approaches in an example
that is easy to conceptualize, we will focus on a simple parametric
model for a single species (derived from fist principles by Allen et al.
2005) as our underlying ``reality''.

@@

Should we include any emprical examples? Using data simulated from a
specified model rather than empirical data permits the comparison
against the true underlying dynamics, setting a bar for the optimal
performance possible.

(Motivation, alternatives, stationarity, examples without a stable node
(limit-cycle models), examples based on observations near a stable node
alone, and why that isn't impossible).

@@

\subsubsection{Discussion of maximum likelihood estimated models}

(MLE models will assume the noise is log-normal, which it is in the
simulation).

Which estimates a Ricker model with $r =$ 1.8501, $K =$ 9.8091, and the
Allen Allele model with $r =$ 2.8079, $K =$ 11.8235 and $C =$ 7.2159.

\subsection{Figure 1:}

\emph{Shows the inferred Gaussian Process compared to the true and
parametric models. Refer to the appendix for details on the GP
posteriors, etc.}

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. (The GP is conditioned on
0,0, shown as a pseudo-data point).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/policies_plot.pdf}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/sim_plot.pdf}
\caption{Gaussian process inference outperforms parametric estimates.
Shown are 100 replicate simulations of the stock dynamics (eq 1) under
the policies derived from each of the estimated models, as well as the
policy based on the exact underlying model.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/posteriors.pdf}
\caption{Histogram of posterior distributions for the estimated Gaussian
Process shown in Figure 1. Prior distributions overlaid.}
\end{figure}

>>>>

\textbf{\emph{Okay, perhaps that's novel, but it's pretty trivial. Isn't
it obvious to everyon that it's trivial?}}

\textbf{\emph{paraphrase advantages of machine learning}}

@@

The essense of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.\\
\textbf{\emph{Figure 1 include curves drawn from the posterior
density?}}

\textbf{\emph{Do we need more specifics on Gaussian process as an
approximation to parametric models? Discussion of Gaussian process vs
other machine learning / forecasting approaches that have less clear
statistical foundations? If so, does this all belong in the
discussion?}}

@@

Once the posterior Gaussian process (GP) has been estimated (e.g.~see
Munch et al. 2005), it is necessary to adapt it in place of the
parametric equation for the stochastic dynamic programming (SDP)
solution (see Mangel and Clark 1988 for a detailed description of
parametric SDP methods) to the optimal policy. The essense of the idea
is straight forward -- we will use the estimated GP in place of the
parametric growth function to determine the stochastic transition matrix
on which the SDP calculations are based. The SDP is solved in a
discritized state space -- both the continuously valued population
densities $X$ and harvest quotas $h$ are first mapped to a bounded,
discrete grid. (For simplicity we will consider a uniform grid, though
for either parametric or GP-based SDP it is often advantageous to use a
non-uniform discretization such as a basis function representation,
e.g.~see (Deisenroth, Rasmussen, and Peters 2009)).

The SDP approach then computes a transition matrix, $\mathbf{F}$. We
demonstrate that calculation is just as straight forward based on the GP
as it is in the classical context using the parametric model. The
${i,j}$ of the transition matrix $F$ entry gives the probability of
transitioning into state $x_i$ given that the system is in state $x_j$
in the previous timestep. To generate the transition matrix based on the
posterior GP, we need only the expected values at each grid point and
the corresponding variances (the diagonal of the covariance matrix), as
shown in Figure 1. Given the mean of the GP posterior at each gridpoint
as the vector $E$ and variance at that point as vector $V$, the
probability of transitioning from state $x_i$ to state $x_j$ is

\[\mathcal{N}\left(x_j \vert  \mu = E_i, \sigma = \sqrt{V_i}\right)\]

where $\mathcal{N}$ is the Normal density at $x_j$ with mean $\mu$ and
variance $\sigma^2$. Strictly speaking, the transition probability
should be calculated by integrating the normal density over the bin of
width $\Delta$ centered at $x_j$. For a sufficiently fine grid that
$f(x_j) \approx f(x_j + \Delta)$, it is sufficient to calculate the
density at $x_j$ and then row-normalize the transition matrix. The
process can then be repeated for each possible discrete value of our
control variable, (harvest $h$).

\textbf{Pseudocode for the determining the transtion matrix from the GP}

@@

This completes the algorithm adapting the GP to the sequential
decision-making problem through SDP, which we believe has not yet been
demonstrated in the literature. We provide an R package implementation
of this, along with the Gaussian process inference, in the supplemental
materials.

@@

Concerns over the potential for tipping points in ecological dynamics
(Scheffer et al. 2001) highlight the dangers of uncertainty in
ecological management and pose a substantial challenge to existing
decision-theoretic approaches (Brozović and Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see @Hughes2013), management is most often concerned with avoiding
potentially catostrophic tipping points before any data is available at
or following a transition that would more clearly reveal these regime
shift dyanmics {[}e.g. @Bestelmeyer2012{]}.

To illustrate the value of the non-parametric Bayesian approach to
management, we focus on example of a system containing such a tipping
point whose dynamics can still be described by a simple, one-dimensional
parametric model.\\We will focus on a simple parametric model for a
single species (derived from fist principles by Allen et al. 2005) as
our underlying ``reality''.

@@

\textbf{\emph{Should we include any emprical examples?}}

\subsubsection{Estimating parametric models}

The maximum likelihood estimates (MLE) will assume that the process
noise $Z_t$ is log-normal, as in the simulation itself. In reality the
noise structure might differ from this assumption, making this a
best-case scenario for the MLE approach.\\The GP must also make an
assumption about the noise structure -- in this case we assume additve
noise, though multiplicative noise would be more accurate. This will
illustrate the flexibility of the GP in fitting different noise
structures even without knowledge of the correct assumptions.

The GP estimate is Bayesian approach which requires specification of
prior distributions for the hyperparameters (see Appendix for details,
or @Rassmussen1996, Munch et al. 2005 for background), as given in Table
1. Then given the data, both the GP and the MLE parametric models can be
estimated as described above, as we implement in R in the code provided
in supplementary materials.

@@

Figure 1 shows the inferred Gaussian Process compared to the true and
the MLE parametric models. The grey shading shows the uncertainty
estimated by the GP. Unlike the MLE estimates, the GP uncertainty
changes depending on where the data is available, allowing for a much
greater level of uncertainty at low stock sizes. While it would be
natural (and straight forward) to condition the GP on passing through
the origin (0,0) (see appendix), the estimate shown here is based only
on the observed data.

\begin{figure}[htbp]
\centering
\includegraphics{figure/gp_plot.png}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance (grey shading)
is also shown. The training data is also shown as black points.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/policies_plot.png}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/sim_plot.png}
\caption{Gaussian process inference outperforms parametric estimates.
Shown are 100 replicate simulations of the stock dynamics (eq 1) under
the policies derived from each of the estimated models, as well as the
policy based on the exact underlying model.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure/posteriors.png}
\caption{Histogram of posterior distributions for the estimated Gaussian
Process shown in Figure 1. Prior distributions overlaid.}
\end{figure}

### 2013-04-08 inst/examples/adp-intro.Rmd

```{r plotting, include=FALSE}
opts_knit$set(upload.fun = socialR::flickr.url)
opts_chunk$set(dev.args=list(bg="transparent"), 
               cache=TRUE, cache.path = "adp/",
               comment=NA, message=FALSE, warning=FALSE)
```

@@

```{r}
a <- 1.5
b <- .5
f <- function(x, z) z * a * x / (x + b)
```

We begin with a simulation method $X_{t+1} = f(X_t, Z_t)$.  For illustration, let us consider $f(X_t, Z_t) = Z_t \frac{a X_t}{b + X_t}$ with a = `r a` and b = `r b`.  We define a statespace $S$

@@

```r
chi <- S
```

@@

The approximate dynamic programming algorithm will perform a finite number $N$ = `r N$ iterations over a window of time $T$ =`r T` in our example.  The algorithm can be described as follows:

@@

```r
Z <- lnorm(T)
```

@@

$$V_t(S_t) = \max_{x_t \in \chi_t} \left(C(S_t, x_t) + \gamma \sum_{s^{\prime} \in mathcal{S}} \mathbb{P}(s^{\prime} | S_t^n, x_t) V_{t+1}^{n-1} s^{\prime} \right)

@@

```{r}
s <- S_0

sapply(0:T, function(t){
  max(
    arg <- sapply(chi, function(x){
      min(s, x) + 
    c(x.n_t = which.max(arg), v.n_t = max(arg))

>>>>

```{r plotting, include=FALSE}
opts_knit$set(upload.fun = socialR::flickr.url)
opts_chunk$set(dev.args=list(bg="transparent"), 
               cache=TRUE, cache.path = "adp/",
               comment=NA, message=FALSE, warning=FALSE,
               tidy=FALSE)
```

@@

``` {r bevholt}
f <- function(x, h, p){
    A <- p[1] 
    B <- p[2] 
    s <- pmax(x-h, 0)
    A * s/(1 + B * s)
}
pars <- c(1.5, 0.05)
K <- (pars[1] - 1)/pars[2]
```

We begin with a simulation method $X_{t+1} = f(X_t, Z_t)$.  For illustration, let us consider $f(X_t, Z_t) = Z_t \frac{a X_t}{b + X_t}$ with a = `r pars[1]` and b = `r pars[2]`.  We define a statespace $S$

@@

```{r}
chi <- S
```

@@

The approximate dynamic programming algorithm will perform a finite number $N$ = `r N` iterations over a window of time $T$ =`r T` in our example.  The algorithm can be described as follows:

@@

```{r}
sigma <- 0.2
Z <- rlnorm(T, 0, sigma)
```

@@

$$V_t(S_t) = \max_{x_t \in \chi_t} \left(C(S_t, x_t) + \gamma \sum_{s^{\prime} \in mathcal{S}} \mathbb{P}(s^{\prime} | S_t^n, x_t) V_{t+1}^{n-1} s^{\prime} \right)$$

@@

```r
s <- S_0
C <- function(S, X) pmin(S, X)

arg <- sapply(chi, function(x) C(s, x) + f(S, x, pars) %*% V)
c(x_nt = which.max(arg), v_nt = max(arg))

## Update V
